---
title: "code_cracking"
output: pdf_document
author: Yongsheng Li
date: "4/5/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Method

The hastings ratio is: $hprob = min(1,\frac{P(D|\theta')\cdot \pi(\theta')q(\theta'\rightarrow{\theta})}{P(D|\theta)\cdot \pi(\theta)q(\theta\rightarrow{\theta'})})$, With our assumptions, $\pi(\theta')q(\theta'\rightarrow{\theta})=\pi(\theta)q(\theta\rightarrow{\theta'})$, so,     
$hprob = min(1,\frac{P(D|\theta')}{P(D|\theta)})=min(1,exp(log(P(D|\theta')-P(D|\theta)))$, and we will move to $\theta'$ if `runif(1)<hprob`.          


1. read in data and normalize the frequency to propabilities

```{r}
# read coded message
input1 = readLines("CodedMessage.txt", n=1000) 

# read frequency table 
q = as.matrix(read.table("LetterPairFreqFrom7Novels.txt",header=F))
# using q as the probabilities, so normalizing to 1 by the sum of all
q=q/sum(q)
colnames(q) = letters
rownames(q) = letters
```

2. extract the word-pairs from the input

```{r}
#
extract_pairs = function(input){
  pairs = c()
  for(i in 1:nchar(input)-1){
    w1 = tolower(substring(input, i, i))
    w2 = tolower(substring(input, i+1, i+1))
    if ((w1 %in% letters) & (w2 %in% letters)){
      pairs = rbind(pairs, c(w1,w2))
    }
  }
  return(pairs)
}
# extract pairs from the long passage
pairs = extract_pairs(input1)
```

3. define a function to calculate log probability given two words and the code table

```{r}
cal_logl = function(w1,w2,q,code){
  w1_real = letters[code==w1]
  w2_real = letters[code==w2]
  prob = q[w1_real, w2_real]
  return(log(prob))
}
```

4. define the mcmc process to sample from the posterior distribution of code table and iterate it for some iterations.

```{r}
mcmc_posterior = function(n,seed=2021){
  set.seed(seed)
  code = sample(letters)
  logl_max = -Inf
  code_max = code
  logl = vector("numeric",n)
  for (i in 1:n){
    new_code = code
    
    # propose a new code table by randomly switching two letters
    proposal = sample(1:26,2)
    temp = new_code[proposal[1]]
    new_code[proposal[1]] = new_code[proposal[2]]
    new_code[proposal[2]] = temp
    
    # calculate the log probability before and after switching
    logl_before = sum(apply(pairs,1,FUN = function(x){cal_logl(x[1],x[2],q,code)}))
    logl_after = sum(apply(pairs,1,FUN=function(x){cal_logl(x[1],x[2],q,new_code)}))
    logl[i] = logl_after
    
    # record the best code table with which maximizes logl_after up to now
    if (logl_after>logl_max){
      logl_max = logl_after
      code_max = new_code
    }
    # define the hprob
    hprob = min(1,exp(logl_after-logl_before))
    
    # move to new code with prob.= hprob
    u = runif(1)
    if (u<hprob){
      code = new_code
    }
    if (i%%1000==0){
      cat(i," iterations completed","\n")
    }
  }
  return(code_max)
}
code = mcmc_posterior(5000, seed=2021)
```


5. decode the coded text using the code sampled from mcmc_posterior.

```{r}
decode = function(code,input){
  text = c()
  for(i in 1:nchar(input)-1){
    w = tolower(substring(input, i, i))
    if (w %in% letters){
      text = c(text,letters[code==w])
    }else{
      text = c(text, w)
    }
  }
  return(paste(text, collapse = ""))
}
text = decode(code,input1)
text=strsplit(text,split=" ")[[1]]
for (i in 1:1087){
  cat(text[i]," ")
  if (i%%10==0){
    cat("\n")
  }
}
```



